---
title: "dotargs"
output: github_document
---

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
<!-- badges: end -->

## A simple framework for building shell interfaces

The purpose of `dotargs` is to significantly reduce the overhead involved in
wrapping shell programs in R. The tools are intended to be intuitive and
lightweight enough to use for data scientists trying to get things done quickly,
but robust and full-fledged enough for developers to extend them to more
advanced use cases.

## Vocabulary to describe operations
Briefly, `dotargs` captures R function arguments (**args**) and converts them to
a vector of commandline **flags**. 

The `dotargs` framework provides three
mechanisms for capturing function arguments:

 - `getDotArgs` captures all arguments passed to `...`
 - `getNamedArgs` captures all keyword arguments defined by the user
 - `getAllArgs` captures both named + dot arguments
 
`argsToFlags` converts the captured arguments to commandline-style flags. This
output can be directly fed to `system2` or `processx`.

Together, they can be used to build user-friendly R interfaces to shell programs
without having to manually implement all commandline flags in R functions.

## Install
```{r, eval=F}
remotes::install_github("snystrom/dotargs")
```

## Examples
```{r}
library(magrittr)
library(dotargs)
```

variables defined in `...` are converted to character vector of flags appropriate for shell commands.
```{r}

myFunction <- function(...){

  flags <- getDotArgs() %>%
    argsToFlags()
  
  return(flags)
}
```

```{r}
myFunction(flag = "var", bool_flag = TRUE)
```

This character vector can be passed to `system2` to build shell commands. 
```{r}
shellCut <- function(text, ...){

  flags <- getDotArgs() %>%
    argsToFlags()

	system2("cut", flags, stdout = T, input = text)

}
```

```{r}
shellCut("hello_world", f = 2, d = "_") 
```

Multiple values can be passed to arguments using vectors
```{r}
shellCut("hello_world_hello", f = c(1,3), d = "_") 
```

### Boolean flags are passed as bool operators

```{r}
shell_ls <- function(dir = ".", ...){
  flags <- getDotArgs() %>% 
    argsToFlags()
  
  system2("ls", c(dir, flags), stdout = T)
}
```

```{r}
shell_ls("R")
```

```{r}
shell_ls("R", l = T)
```

### Named vectors can be used to provide user-friendly aliases for single-letter flags
```{r}

shell_ls_alias <- function(dir = ".", ...){
  
  argsDict <- c("long" = "l")
  
  flags <- getDotArgs() %>% 
    argsToFlags(argsDict)
  
  system2("ls", c(dir, flags), stdout = T)
}
```

```{r}
shell_ls_alias("R", long = T)
```

```{r}
shellCut_alias <- function(text, ...){

  argsDict <- c("sep" = "d",
                "fields" = "f")
    
	flags <- getDotArgs() %>%
		argsToFlags(argsDict)

	system2("cut", flags, stdout = T, input = text)
}
```

```{r}
shellCut_alias("hello_world", fields = 2, sep = "_") 
```

## Abstraction of command path handling

A common pattern when designing shell interfaces is to ask the user to give an
absolute path to the target shell utility. It is common to pass this information
from the user to R by using either R environment variables defined in `.Renviron`,
using options (set with `option()`, and got with `getOption()`), having the user
explicitly pass the path in the function call, or failing this, using a default
install path.

`build_path_handler()` is a macro which returns a function that returns a valid
path to the target by heirarchically searching a series of possible locations.

For example, to build an interface to the "MEME" suite, which is by default
installed to `~/meme/bin`, one could build the following:

```{r}
handle_meme_path <- build_path_handler(default_path = "~/meme/bin")

handle_meme_path()
```

To only search the R environment variable "MEME_PATH", one could build:
```{r}
handle_meme_path <- build_path_handler(environment_var = "MEME_PATH")
```


```{r, error = T}
# Without environment variable defined
handle_meme_path()
```

```{r}
# With environment varialbe defined
Sys.setenv("MEME_PATH" = "~/meme/bin")
handle_meme_path()
```

Multiple arguments can be used, and they will be searched from most-specific, to most-general. 
```{r}
handle_meme_path <- build_path_handler(environment_var = "MEME_PATH",
                                       default_path = "~/meme/bin")
```

For example, if "MEME_PATH" is invalid on my machine, the handler will return
the default path as long as the default is also valid on my machine.
```{r}
Sys.setenv("MEME_PATH" = "bad/path")
handle_meme_path()
```

### Support for tool utilities
Some software, like the MEME suite is distributed as several binaries located in
a common directory. To allow interface builders to officially support specific binaries, 
each binary can be defined as a "utility" within the build path.

Here, I will include two tools from the MEME suite, AME, and DREME (distributed
as binaries named "ame", and "dreme").
```{r}
handle_meme_path <- build_path_handler(environment_var = "MEME_PATH",
                                       default_path = "~/meme/bin",
                                       utils = c("dreme", "ame"))
```

handler functions have two optional arguments: `path` and `util`. `path` acts as
an override to the defaults provided when building the handler. User-provided
path variables will always be used instead of provided defaults. This is to
catch problems from the user and not cause unexpected user-level behavior.
```{r, error = T}
handle_meme_path("bad/path")
```

`util` specifies which utility path to return (if any). The path handler will
throw an error if the utility is not found in any of the specified locations.
```{r, error = T}
handle_meme_path(util = "dreme")
```

## Bringing it all together
Using a `get*Args()` family function to get and convert function arguments to
commandline flags. The path handler returns the correct `command` call which can
be passed to `system2` or `processx` along with the flags generated from `argsToFlags`.

This makes for a robust shell wrapper without excess overhead.

In the `runDreme` function below, the user can pass any valid `dreme` argument
using the rules for command args defined above to `...`. Allowing `meme_path` as
a function argument and passing it to `handle_meme_path` allows the user to
override the default search path which is: `MEME_PATH` environment variable,
followed by the `~/meme/bin` default install.

```{r}
handle_meme_path <- build_path_handler(environment_var = "MEME_PATH",
                                       default_path = "~/meme/bin",
                                       utils = c("dreme", "ame"))

runDreme <- function(..., meme_path = NULL){
  flags <- getDotArgs() %>% 
    argsToFlags()
  
  command <- handle_meme_path(path = meme_path, util = "dreme")
  
  system2(command, flags)
}
```

Commands can now run through `runDreme` by passing flags as function arguments.
```{r, error = T}
runDreme(h = T)
```

## Unsafe operations

**WARNING:** It's still possible to do unsafe operations as follows, so please be careful how you build system calls.

```{r}
shellCut_unsafe <- function(text, ...){

  flags <- getDotArgs() %>%
    argsToFlags()

	system2("echo", c(text , "|", "cut", flags), stdout = T)

}

shellCut_unsafe("hello_world", f = 2, d = "_ && echo test")
```

**NOTE** even if when setting `stdout = TRUE` the second command doesn't appear
in the output, it will still have run. 

A more extreme example of what can happen is here, where `~/deleteme.txt` will be removed silently.

I promise I'll get around to sanitizing user input eventually.
```{r, eval=F}
shellCut("hello_world", f = 2, d = "_ && rm ~/deleteme.txt")
```
