---
title: "dotargs"
output: github_document
---

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
<!-- badges: end -->

## A simple framework for building shell interfaces

`dotargs` provides two main functions: `getDotArgs` and `argsToFlags` which can act
as a useful backend for converting keyword arguments passed to `...` in a function
call into a vector of shell command flags which can be passed to `system2` or
`processx`. Together, they can be used to build user-friendly R interfaces to
shell programs without having to manually implement all commandline flags in R
functions.

Also `getAllArgs` and `getNamedArgs` can be used to get all named arguments +
dots, or named arguments only (no dots), depending on your use-case.

## Install
```{r, eval=F}
remotes::install_github("snystrom/dotargs")
```

## Examples
```{r}
library(magrittr)
library(dotargs)
```

variables defined in `...` are converted to character vector of flags appropriate for shell commands.
```{r}

myFunction <- function(...){

  flags <- getDotArgs() %>%
    argsToFlags()
  
  return(flags)
}
```

```{r}
myFunction(flag = "var", bool_flag = TRUE)
```

This character vector can be passed to `system2` to build shell commands. 
```{r}
shellCut <- function(text, ...){

  flags <- getDotArgs() %>%
    argsToFlags()

	system2("cut", flags, stdout = T, input = text)

}
```

```{r}
shellCut("hello_world", f = 2, d = "_") 
```

Multiple values can be passed to arguments using vectors
```{r}
shellCut("hello_world_hello", f = c(1,3), d = "_") 
```

### Boolean flags are passed as bool operators
```{r}
shell_ls <- function(dir = ".", ...){
  flags <- getDotArgs() %>% 
    argsToFlags()
  
  system2("ls", c(dir, flags), stdout = T)
}
```

```{r}
shell_ls("R")
```

```{r}
shell_ls("R", l = T)
```

### Named vectors can be used to provide user-friendly aliases for single-letter flags
```{r}

shell_ls_alias <- function(dir = ".", ...){
  
  argsDict <- c("long" = "l")
  
  flags <- getDotArgs() %>% 
    argsToFlags(argsDict)
  
  system2("ls", c(dir, flags), stdout = T)
}
```

```{r}
shell_ls_alias("R", long = T)
```

```{r}
shellCut_alias <- function(text, ...){

  argsDict <- c("sep" = "d")
    
	flags <- getDotArgs() %>%
		argsToFlags(argsDict)

	system2("cut", flags, stdout = T, input = text)
}
```

```{r}
shellCut_alias("hello_world", f = 2, sep = "_") 
```

## Unsafe operations

**WARNING:** It's still possible to do unsafe operations as follows, so please be careful how you build system calls.

```{r}
shellCut_unsafe <- function(text, ...){

  flags <- getDotArgs() %>%
    argsToFlags()

	system2("echo", c(text , "|", "cut", flags), stdout = T)

}

shellCut_unsafe("hello_world", f = 2, d = "_ && echo test")
```

**NOTE** even if when setting `stdout = TRUE` the second command doesn't appear
in the output, it will still have run. 

A more extreme example of what can happen is here, where `~/deleteme.txt` will be removed silently.

I promise I'll get around to sanitizing user input eventually.
```{r, eval=F}
shellCut("hello_world", f = 2, d = "_ && rm ~/deleteme.txt")
```
