---
title: "cmdfun"
output: github_document
---

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![R build status](https://github.com/snystrom/cmdfun/workflows/R-CMD-check/badge.svg)](https://github.com/snystrom/cmdfun/actions)
[![Codecov test coverage](https://codecov.io/gh/snystrom/cmdfun/branch/master/graph/badge.svg)](https://codecov.io/gh/snystrom/cmdfun?branch=master)
<!-- badges: end -->

## A simple framework for building shell interfaces

The purpose of `cmdfun` is to significantly reduce the overhead involved in
wrapping shell programs in R. The tools are intended to be intuitive and
lightweight enough to use for data scientists trying to get things done quickly,
but robust and full-fledged enough for developers to extend them to more
advanced use cases.

## Installation

Install the development version of `cmdfun` with:
```{r, eval=F}
if (!requireNamespace("remotes"))
  install.packages("remotes")

remotes::install_github("snystrom/cmdfun")
```

## Quick Examples

`cmdfun` attempts to solve the problem of wrapping external software in R. 
Calling external software is done with `system2` or `processx`.

For example, calling `ls -l`.
```{r, eval=F}
system2("ls", "-l", stdout = TRUE)
```
```{r, echo = F}
system2("ls", "-l | tail -n5", stdout = TRUE)
```

However, when using multiple commandline flags each flag must be passed as a
member of a character vector as follows:

When calling `ls -l -i`
```{r, eval=FALSE}
system2("ls", c("-l", "-i"), stdout = TRUE)
```
```{r, echo = FALSE}
system2("ls", c("-l", "-i | tail -n5"), stdout = TRUE)
```

This becomes even more difficult if trying to support user input, as a
significant amount of overhead is required to parse user inputs and optional
flags into these vectors.

`cmdfun` provides utilities for converting **function arguments** into **lists**
which can easily convert to **character vectors** suitable for use with
`system2` or `processx`.

```{r}
library(cmdfun)

myFunction <- function(input, option1){
  # Grabs named arguments as key/value pairs
  cmd_args_named()
}

(argsList <- myFunction("myInput.txt", "foo"))
```
```{r}
# Converts list to character vector of flags & values
cmd_list_to_flags(argsList)
```

Additionally, `cmdfun` provides utilites for searching & checking valid tool
installs, expecting system behavior, and helpful error handling to allow simple
construction of external tool wrappers.

## More Details
See [https://snystrom.github.io/cmdfun/articles/cmdfun.html](https://snystrom.github.io/cmdfun/articles/cmdfun.html)
for the most recent documentation and to learn about all `cmdfun` features.

To file bug reports, please visit
[https://github.com/snystrom/cmdfun/issues](https://github.com/snystrom/cmdfun/issues)
while providing a [reproducible example](https://reprex.tidyverse.org/) of your
issue.