---
title: "Checking valid arguments"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{help_checking}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
# This will only build on unix
is_unix <- .Platform$OS.type == "unix"
knitr::opts_chunk$set(eval = is_unix)
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(cmdfun)
library(magrittr)
```

```{r}
out <- processx::run("cut", "--help", error_on_status = FALSE)
out$stdout %>% 
  cmd_help_parse_flags(split_newline = TRUE)
```


# OLD

Methods to get help lines into R. Examples with processx and system2 (I suggest `processx` because it's handling is better, although it needs more steps)
```{r}
# with processx
o <- processx::run("tar", "--help", error_on_status = FALSE)
fn_flags <- cmd_help_parse_flags(o$stdout, split_newline = TRUE)

# with system2
lines <- system2("tar", "--help", stdout = T) 
fn_flags <- cmd_help_parse_flags(lines)
fn_flags %>% head(10)
  gsub(", -", "\n-", .) %>% 
  gsub(",", "", .) %>% 
  gsub("=.+", "", .) %>% 
  gsub("\\[", "", .) %>% 
  unique

# Custom function to clean up final parsed flags
parse_tar_help <- function(lines){
  lines %>% 
    gsub(", -", "\n-", .) %>% 
    gsub(",", "", .) %>% 
    gsub("=.+", "", .) %>% 
    gsub("\\[", "", .) %>% 
    unique
}
```

```{r}
tar <- function(...){
  cmd_args_dots() %>% 
    cmd_list_to_flags()
}

tar_withCheck <- function(...){
  tarHelp <- system2("tar", "--help", stdout = T) 
  
  tarflags <- tarHelp %>% 
    cmd_help_parse_flags() %>% 
    parse_tar_help()
  
  inputFlags <- cmd_args_dots() %>% 
    cmd_list_interp() 
    #purrr::set_names(~{gsub("_", "-", .)})
  
  # check that inputs are valid
  cmd_help_flags_similar(tarflags, names(inputFlags), ~{gsub("-", "_", .)}) %>% 
    cmd_help_flags_suggest()
}

tar_withCheck(utz = "value", exclude_caces = TRUE)
tar_withCheck(c = T, concatenate = T)
```

help checking is expensive (relative to not doing it), so it makes sense to only do it if there's an error. 

We can write a simple function to check 
```{r}
check_tar_flags <- function(flags){
  flag_names <- names(flags)
  
  tarhelp_out <- processx::run("tar", "--help", error_on_status = F)
  
  tarhelp_out$stdout %>% 
    cmd_help_parse_flags(split_newline = TRUE) %>% 
    cmd_help_flags_similar(flag_names,  ~{
      gsub("-", "_", .) %>% 
        gsub(",", "", .)
      }) %>% 
    cmd_help_flags_suggest()
}
```
```{r}
check_tar_flags(list("utd" = 1, "verzion" = 2))
```

This is where `processx` shines, but error checking can be done using `system2`
as well. We catch errors by checking the exit status, where non-zero exit status
is considered an error.
```{r}
tar_withCheck <- function(file, ...){
  
  inputFlags <- cmd_args_dots() %>% 
    cmd_list_interp() %>% 
    purrr::set_names(~{gsub("_", "-", .)})
  
  flags <- inputFlags %>% 
    cmd_list_to_flags()
  
  flags <- c(flags, file)
  ps_out <- processx::run("tar", flags, error_on_status = F)
  
  if (ps_out$status != 0){
    # Print stdout & stderr to user
    message(ps_out$stdout)
    message(ps_out$stderror)
    # Suggest flags
    check_tar_flags(inputFlags)
  }
  
}
```


```{r}
tar_withCheck("test.txt", uts = "value") 
```

