---
title: "Writing shell wrappers with cmdfun"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Writing shell wrappers with cmdfun}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

old_path <- Sys.getenv("PATH")
Sys.setenv(PATH = paste(old_path, "/bin", sep = ":"))
```

## A simple framework for building shell interfaces

The purpose of `cmdfun` is to significantly reduce the overhead involved in
wrapping shell programs in R. The tools are intended to be intuitive and
lightweight enough to use for data scientists trying to get things done quickly,
but robust and full-fledged enough for developers to extend them to more
advanced use cases.

Briefly, `cmdfun` captures R function arguments (**args**) as a base R `list`
and converts them to a vector of commandline **flags**.

## Grabbing function arguments as lists

The `cmdfun` framework provides three
mechanisms for capturing function arguments:

 - `cmd_args_dots()` captures all arguments passed to `...`
 - `cmd_args_named()` captures all keyword arguments defined by the user
 - `cmd_args_all()` captures both named + dot arguments
 
`cmd_list_interp` converts the captured argument list to a parsed list of
flag/value pairs (details below). This output can be useful for additional
handling of special flag assignments from within R.

`cmd_list_to_flags` converts a list to a vector of
commandline-style flags using the list names as flag names and the list values
as the flag values (empty values return only the flag). This output can be
directly fed to `system2` or `processx`.

Together, they can be used to build user-friendly R interfaces to shell programs
without having to manually implement all commandline flags in R functions.

```{r}
library(magrittr)
library(cmdfun)
```

The `cmd_args` family of functions operate **within** a function environment to
capture the arguments. Therefore, to examine their behavior, they must be
wrapped in functions.

Here I'll compare the differences between the three `cmd_args` functions.
```{r}
get_all <- function(arg1, arg2, ...){
  cmd_args_all()
}

get_named <- function(arg1, arg2, ...){
  cmd_args_named()
}

get_dots <- function(arg1, arg2, ...){
  cmd_args_dots()
}
```

```{r}
# cmd_args_all() gets all keword arguments and arguments passed to "..."
(argsListAll <- get_all("input", NA, bool = TRUE, vals = c(1,2,3)))
```

```{r}
# cmd_args_named() gets all keword arguments, excluding arguments passed to "..."
(argsListNamed <- get_named("input", NA, bool = TRUE, vals = c(1,2,3)))
```

```{r}
# cmd_args_dots() gets all arguments passed to "...", excluding keyword arguments
(argsListDots <- get_dots("input", NA, bool = TRUE, vals = c(1,2,3)))
```

The captured argument lists contain the argument names as list names, and the
argument values as list values.

## Create R representations of commandline flags

Passing flags to commandline software inherently varies from how arguments are
passed to R functions. For example, some flags require values to follow them,
while others do not (ie `head -n 5` vs `ls -l`). In some cases, flags can have
multiple values assigned to them, passed as comma-separated entries (ie `cut -f 1,2,3`). 

To facilitate the conversion of these concepts between R and the shell,
`cmd_list_interp` **interprets** R data structures in each list entry and
converts them where necessary to prepare them for final conversion to a
character vector.

Argument values are interpreted according to the following rules:

| Argument Value Type  | `cmd_list_interp` behavior |
|:--------------------:|:--------------------------:|
| `character(1)`       | keep                       |
| `numeric(1)`         | keep                       |
| `character` vector   | keep                       |
| `numeric` vector     | keep                       |
| `factor`             | keep                       |
| `TRUE`               | Convert to: ""             |
| `FALSE`              | Drop entry from list       |
| `NA`                 | Drop entry from list       |
| `NULL`               | Drop entry from list       |


```{r}
# Note that `arg2` is dropped, and `bool` is converted to ""
cmd_list_interp(argsListAll) 
```

Commandline tools can have hundreds of arguments, many with uninformative, often
single-letter, names. To prevent developers from having to write aliased
function arguments for all, often conflicting flags, `cmd_list_interp` can
additionally use a lookup table to allow developers to provide informative
function argument names for unintuitive flags.

```{r}
(flagList <- cmd_list_interp(argsListAll, c("bool" = "b")))
```

## Convert lists to flag vectors

After interpreting argument lists with `cmd_list_interp`, the resulting list can
be coerced into a vector suitable for passing to `system2` or `processx` using
`cmd_list_to_flags`. `cmd_list_to_flags` will produce the following vector
values for each name/value pair in the list:

| Argument Value Type  | `cmd_list_to_flags` behavior |
|:--------------------:|:----------------------------:|
| `character(1)`       | `c("-name", "value")`        |
| `numeric(1)`         | `c("-name", "value")`        |
| `character` vector   | `c("-name", "a,b,c")`        |
| `numeric` vector     | `c("-name", "1,2,3")`        |
| `factor`             | same as `character`          |
| empty string: ""     | `c("-name")`                 |

```{r}
cmd_list_to_flags(flagList)
```

## Examples using unix tools

Here are two examples wrapping common shell utilities `ls` and `cut`.

### Wrapping `ls` with cmdfun

These tools can be used to easily wrap `ls`, a command which lists files in the target directory.
```{r}
library(magrittr)

shell_ls <- function(dir = ".", ...){
  # grab arguments passed to "..." in a list
  flags <- cmd_args_dots() %>% 
    # prepare list for conversion to vector
    cmd_list_interp() %>% 
    # Convert the list to a flag vector
    cmd_list_to_flags()
  
  # Run ls shell command
  system2("ls", c(flags, dir), stdout = TRUE)
}
```

```{r}
# list all .md files in ../
shell_ls("../*.md")
```

#### Boolean flags are passed as bool operators
`ls -l` can be mimiced by passing `l = TRUE` to '...'.
```{r}
shell_ls("../*.md", l = TRUE)
```

#### Using a lookup table to make user-friendly argument names

For example, allowing `long` to act as `-l` in `ls`.
```{r}

shell_ls_alias <- function(dir = ".", ...){
  
  # Named vector acts as lookup table
  # name = function argument
  # value = flag name
  names_arg_to_flag <- c("long" = "l")
  
  flags <- cmd_args_dots() %>% 
    # Use lookup table to manage renames
    cmd_list_interp(names_arg_to_flag) %>% 
    cmd_list_to_flags()
  
  system2("ls", c(flags, dir), stdout = TRUE)
}
```

```{r}
shell_ls_alias("../*.md", long = TRUE)
```

### Wrapping `cut` with cmdfun
Here is another example wrapping `cut` which separates text on a delimiter (set
with `-d`) and returns selected fields (set with `-f`) from the separation.
Again, we use a lookup table to create the optional `sep` and `fields` arguments
which specify `-d` and `-f`, respectively.
```{r}
shell_cut <- function(text, ...){

  names_arg_to_flag <- c("sep" = "d",
                         "fields" = "f")
    
	flags <- cmd_args_dots() %>%
		cmd_list_interp(names_arg_to_flag) %>% 
	  cmd_list_to_flags()

	system2("cut", flags, stdout = T, input = text)
}
```

```{r}
shell_cut("hello_world", fields = 2, sep = "_") 
```

Multiple values can be passed to arguments using vectors
```{r}
# Note that the flag name values are accepted even when using a lookup table
shell_cut("hello_world_hello", f = c(1,3), d = "_") 
```

## Abstraction of command path detection

Command executables can be stored in different locations across devices,
therefore another barrier to wrapping external software is detecting the
location of the desired tool. The simplest solution is to ask the user to
provide the location to their install, however, requiring this for every function
call can become repetitive and clunky. To reduce this cognitive overhead, a
common pattern when designing shell interfaces is to ask the user to pass this
information to R by using either R environment variables defined in `.Renviron`,
using options (set with `options()`, and got with `getOption()`). Fallback
options can include having the user explicitly pass the path each time in the
function call, or failing this, using a default install path.

`cmd_path_search()` is a macro which returns a function that returns a valid
path to the target by hierarchically searching a series of possible locations
according to the following hierarchy:

1. Manually passing the install path to the function call
2. An option set using `options(option_name = "value")`
3. An environment variable set in `.Renviron`
4. A default install location

The resulting search function will always prefer the most specific option (ie
when a user explicitly passes a path) before falling back to a less-specific
assignment. It is up to the designer whether to support any or all of these
features.

For example, to build an interface to the "MEME" suite, which is by default
installed to `~/meme/bin`, one could build the following:

```{r, echo = FALSE}
# This handles build system not having meme installed
# creates empty files representing real install
search_meme_path <- cmd_path_search(default_path = "~/meme/bin", utils = c("ame", "dreme"))
meme_is_installed <- cmd_install_is_valid(search_meme_path)

dummy_meme <- FALSE
if (!meme_is_installed()) {
  meme_loc <- "~/meme/bin"
  dir.create(meme_loc, recursive = TRUE)
  file.create(paste(meme_loc, "ame", sep = "/"))
  file.create(paste(meme_loc, "dreme", sep = "/"))
  dummy_meme <- TRUE
}

```

This will search for `~/meme/bin` and either return a valid path if it exists,
or throw an error if it can't be found.
```{r}
search_meme_path <- cmd_path_search(default_path = "~/meme/bin")

search_meme_path()
```

The user can always pass their own path which will override the default location. If this path is invalid, the search function will error.
```{r, error = TRUE}
search_meme_path("bad/path")
```

To instead only search the R environment variable "MEME_PATH", one could build:
```{r}
search_meme_path <- cmd_path_search(environment_var = "MEME_PATH")
```


```{r, error = T}
# Without environment variable defined
search_meme_path()
```

```{r}
# With environment variable defined
Sys.setenv("MEME_PATH" = "~/meme/bin")
search_meme_path()
```

Multiple arguments can be used, and they will be searched from most-specific, to most-general. 
```{r}
search_meme_path <- cmd_path_search(environment_var = "MEME_PATH",
                                       default_path = "~/meme/bin")
```

For example, if "MEME_PATH" is invalid on my machine, the search_function will return
the default path as long as the default is also valid on my machine.
```{r}
Sys.setenv("MEME_PATH" = "bad/path")
search_meme_path()
```

As always, if the user passes their own path, this will take precedence.
```{r, error = TRUE}
search_meme_path(path = "bad/path")
```

### Support for tool utilities
Some software, like the MEME suite is distributed as several binaries located in
a common directory. To allow interface builders to officially support specific binaries, 
each binary can be defined as a "utility" within the build path.

Here, I will include two tools from the MEME suite, AME, and DREME (distributed
as binaries named "ame", and "dreme"). The user can set the binary location by
setting the `MEME_PATH` environment variable, passing their own path, or fall
back to the default install location.
```{r}
search_meme_path <- cmd_path_search(environment_var = "MEME_PATH",
                                       default_path = "~/meme/bin",
                                       utils = c("dreme", "ame"))
```

search_function functions have two optional arguments: `path` and `util`. `path` acts as
an override to the defaults provided when building the search_function. User-provided
path variables will always be used instead of provided defaults. This is to
catch problems from the user and not cause unexpected user-level behavior.
```{r, error = T}
search_meme_path("bad/path")
```

`util` specifies which utility path to return (if any). The path search_function will
throw an error if the utility is not found in any of the specified locations.
```{r, error = T}
search_meme_path(util = "dreme")
```

The `cmd_install_check` function can be lightly wrapped by package builders to
verify and print a user-friendly series of checks for a valid tool install. it
takes as input the output of `cmd_path_search` and an optional user-override
`path`. The search logic is inherited from the path search function, so the
options and environment variables are also searched.

Here I build a function for checking a users `meme` install.
```{r}
check_meme_install <- function(path = NULL){
  cmd_install_check(search_meme_path, path = path)
}
```


```{r}
# searches default meme search locations
check_meme_install()
```

```{r}
# uses user override
check_meme_install('bad/path')
```

If you want to write your own install checker instead of using the
`cmd_install_check` function, `cmdfun` also provides the `cmd_ui_file_exists` function
for printing pretty status messages.
```{r}
cmd_ui_file_exists("bad/file")
cmd_ui_file_exists("~/meme/bin")
```

### Internal install validators
`cmdfun` also provides a macro `cmd_install_is_valid()` to construct
functions returning boolean values testing for an install path. These are useful
in function logic, or package development for setting conditional examples or
function hooks that depend on a command install. `cmd_install_is_valid()`
takes a path search function as input, so any `options`, `.Renviron`, or
default install location logic propagates to these functions as well.

```{r}
meme_installed <- cmd_install_is_valid(search_meme_path)
meme_installed()
```

This also works on utils defined during path search construction.
```{r}
ame_installed <- cmd_install_is_valid(search_meme_path, util = "ame")
ame_installed()
```


## Bringing it all together
Using a `cmd_args_` family function to get and convert function arguments to
commandline flags. The path search function returns the correct `command` call which can
be passed to `system2` or `processx` along with the flags generated from `cmd_list_to_flags`.

This makes for a robust shell wrapper without excess overhead.

In the `runDreme` function below, the user can pass any valid `dreme` argument
using the rules for command args defined above to `...`. Allowing `meme_path` as
a function argument and passing it to `search_meme_path` allows the user to
override the default search path which is: the `MEME_PATH` environment variable,
followed by the `~/meme/bin` default install.

```{r}
search_meme_path <- cmd_path_search(environment_var = "MEME_PATH",
                                       default_path = "~/meme/bin",
                                       utils = c("dreme", "ame"))

runDreme <- function(..., meme_path = NULL){
  flags <- cmd_args_dots() %>% 
    cmd_list_interp() %>% 
    cmd_list_to_flags()
  
  dreme_path <- search_meme_path(path = meme_path, util = "dreme")
  
  system2(dreme_path, flags)
}
```

Commands can now run through `runDreme` by passing flags as function arguments.
```{r, eval=FALSE}
runDreme(version = TRUE)
```
```{r, echo = FALSE}
cat("5.1.1")
```

If users have issues with the install, they can run `check_meme_install()` to verify the tools are being detected by R.

# Additional Features
## Restrict argument matching
each `cmd_args_` family function accepts a character vector of names to `keep` or
`drop` arguments which will restrict command argument matches to values in
`keep` (or ignore those in `drop`). As of now, `keep` and `drop` are mutually
exclusive.

This can be useful to allow only some function arguments to be captured as
flags, while others can be used for function logic.

```{r}
myFunction <- function(arg1, arg2, someText = "default"){
  flags <- cmd_args_named(keep = c("arg1", "arg2")) %>% 
    cmd_list_interp() %>% 
    cmd_list_to_flags()
  
  print(someText)
  
  return(flags)
}

myFunction(arg1 = "blah", arg2 = "blah")
```


```{r}
myFunction(arg1 = "blah", arg2 = "blah", someText = "hello world")
```

## Manipulating list objects
For the most part, the [purrr](https://purrr.tidyverse.org/) library is the most
useful toolkit for operations on list objects.

`cmdfun` provides additional helper functions to handle common manipulations.

`cmd_list_drop` operates on argument & flag lists to drop all entries corresponding to a
certain name, specific name/value pairs, or by index position. Conversely,
`cmd_list_keep` functions identically but for keeping entries.

```{r}
myList <- list('value1' = TRUE,
               'value2' = "Hello",
               'value2' = 1:4)

cmd_list_keep(myList, "value2")
```

```{r}
cmd_list_keep(myList, c("value2" = "Hello"))
```

```{r}
cmd_list_drop(myList, "value2")
```

These functions can be useful for ignoring setting certain flags if the user set
them to a specific value.

```{r}
myFunction <- function(arg1, arg2){
  flags <- cmd_args_named() %>% 
    cmd_list_interp() %>% 
    # if arg2 == "baz", don't include it 
    cmd_list_drop(c("arg2" = "baz")) %>% 
    cmd_list_to_flags()
  
  return(flags)
}

myFunction(arg1 = "foo", arg2 = "bar")
myFunction(arg1 = "foo", arg2 = "baz")
```

## Expecting output files
Sometimes a commandline function returns multiple output files you want to check for after the run.

`cmd_error_if_missing` accepts a vector or list of files & checks that they exist.

`cmdfun` additionally provides a few convenience functions for generating lists
of expected files. `cmd_file_combn` generates combinations of extension/prefix
file names. The output can be passed to `cmd_error_if_missing` which will error if
a file isn't found on the filesystem.
```{r}
cmd_file_combn(ext = c("txt", "xml"), prefix = "outFile")
```

```{r}
cmd_file_combn(ext = "txt", prefix = c("outFile", "outFile2", "outFile3"))
```

Alternately, `cmd_file_expect` will build a list of expected files and check whether they exist. This is a wrapper around `cmd_file_combn %>% cmd_error_if_missing`.

## Error checking user input

When using `cmdfun` to write lazy shell wrappers, the user can easily mistype a
commandline flag since there is not text completion. Some programs behave
unexpectedly when flags are typed incorrectly, and for this reason return
uninformative error messages. `cmdfun` has built-in methods to automatically
populate a list of valid flags from a command's help-text.

Alternatively, package builders could pass a vector of allowed flag names to
check against if they didn't want to parse help text. The goal is maximum flexibility.

The following example demonstrates how to parse help text (in this case from
`tar`) into a vector of allowed flags. This vector is compared to the user-input
flags (`user_input_flags` below), and tries to identify misspelled function
arguments.

Here, the user has accidentally used the argument `delte` instead of `delete`.
`cmdfun` tries to be helpful and identify the misspelling for the user.
```{r, error=T}
user_input_flags <- c("delte")

system2("tar", "--help", stdout = TRUE) %>% 
  cmd_help_parse_flags() %>% 
  # Compares User-input flags to parsed commandline flags
  # returns flags that match based on edit distance
  cmd_help_flags_similar(user_input_flags) %>% 
  # Prints error message suggesting the most similar flag name
  cmd_help_flags_suggest()
```


## Unsafe operations

**WARNING:** It's still possible to do unsafe operations as follows, so please be careful how you build system calls.

```{r}
shellCut_unsafe <- function(text, ...){

  flags <- cmd_args_dots() %>%
    cmd_list_interp() %>% 
    cmd_list_to_flags()

	system2("echo", c(text , "|", "cut", flags), stdout = TRUE)

}

shellCut_unsafe("hello_world", f = 2, d = "_ && echo unsafe operation!")
```

**NOTE** even if when setting `stdout = TRUE` the second command doesn't appear
in the output, it will still have run. 

A more extreme example of what can happen is here, where `~/deleteme.txt` will be removed silently.

I promise I'll get around to sanitizing user input eventually, I am still
reasoning about the best way to do this. You can provide feedback on this process at [this issue](https://github.com/snystrom/cmdfun/issues/6).
```{r, eval=F}
shellCut("hello_world", f = 2, d = "_ && rm ~/deleteme.txt")
```

```{r teardown, echo = F}
if (dummy_meme) {
  unlink(meme_loc, recursive = TRUE)
}
```

